# National Emissions Inventory (NEI)

 
```{r}
# set up working directory
setwd('/home/daria/Courses/R/Coursera/EDA/Week3/exdata-data-NEI_data')
library(dplyr)
library(ggplot2)

# read two files
NEI <- readRDS("summarySCC_PM25.rds")
SCC <- readRDS("Source_Classification_Code.rds")
SCC1 <- select(SCC, SCC, Short.Name, SCC.Level.Two)
```


## First look at data 
```{r}
# merging data
data <- merge(NEI, SCC1, by = "SCC")

head(data)
names(data)
```


## How total emissions from PM2.5 decreased in the United States from 1999 to 2008? 
```{r}
# group data by year
by_year <- group_by(data, year)

# calculate sum of emissions for each year
by_year_total <- summarize(by_year, total_emission = sum(Emissions))
by_year_total

# create plot
ggplot(data = by_year_total, aes(x = factor(year), y = total_emission,)) +
 geom_bar(stat='identity', fill = 'red')+
  ggtitle('PM2.5 Emission in USA in 1999 - 2008')+
  xlab('Year') +
  ylab('Total Emission')

# save plot
ggsave(file = 'plot1.png')
```

## How total emissions from PM2.5 decreased in the Baltimore City, Maryland (fips == "24510") from 1999 to 2008? 
```{r}
# filtering grouped by year data by fip of Baltimore City
balt_by_year <- filter(by_year, fips == '24510')

# get the sum of emission for Baltimore for each year
balt_by_year_total <- summarize(balt_by_year, total_emission = sum(Emissions))
balt_by_year_total

# create plot
ggplot(data = balt_by_year_total, aes(x = factor(year), y = total_emission,)) +
 geom_bar(stat='identity', fill = 'blue')+
  ggtitle('PM2.5 Emission in the Baltimore City in 1999 - 2008')+
  xlab('Year') +
  ylab('Total Emission')

#save plot
 ggsave(file = 'plot2.png')
```


## Of the four types of sources indicated by the type (point, nonpoint, onroad, nonroad) variable, which of these four sources have seen decreases in emissions from 1999–2008 for Baltimore City? Which have seen increases in emissions from 1999–2008? 
```{r}
# filter merged data by fip of Baltimore
data_balt <- filter(data, fips == '24510')

# group it by year and type
by_year_type <- group_by(data_balt, year, type)

# sum emissions for each year and year type
by_year_type_sum <- summarise(by_year_type, emission = sum(Emissions))

# ungroup dataframe before plotting
by_year_type_sum <- ungroup(by_year_type_sum)
by_year_type_sum

# create plot
ggplot(data = by_year_type_sum, aes(x = year, y = as.numeric(emission))) +
 geom_line(aes(color = type), size = 4)+
  ggtitle('PM2.5 Emission in the Baltimore in 1999 - 2008')+
  xlab('Year') +
  ylab('Total Emission')

# save plot
ggsave(file = 'plot3.png')
```

## Across the United States, how have emissions from coal combustion-related sources changed from 1999–2008?
```{r}
# get the observations with combustion and coal only
comb <- data[grep('[cC]omb',data$Short.Name),]
comb_coal <- comb[grep('[cC]oal',comb$Short.Name),]

# group data by year and find sum of emissions by each year
by_year <- group_by(comb_coal, year)
by_year_total <- summarize(by_year, total_emission = sum(Emissions))
by_year_total

#create plot
ggplot(data = by_year_total , aes(x = factor(year), y = total_emission)) +
 geom_bar(fill =I('#FF8B00'), stat = 'identity')+
  ggtitle('Emissions from coal combustion-related sources in 1999–2008')+
  xlab('Year') +
  ylab('Total Emission')

# save plot
 ggsave(file = 'plot4.png')
```

## How have emissions from motor vehicle sources changed from 1999–2008 in Baltimore City?
```{r}
# filter data for Baltimore
balt <- filter(data, fips == 24510)

# get data for emissions from motor vehicle sources
vehicle <- balt[grep('[vV]ehicle',balt$SCC.Level.Two),]

# group by year and calculate total emission for each year
by_year <- group_by(vehicle, year)
by_year_total <- summarize(by_year, total_emission = sum(Emissions))

# create plot
ggplot(data = by_year_total , aes(x = factor(year), y = total_emission)) +
 geom_bar(fill =I('#66CD00'), stat = 'identity')+
  ggtitle('Emissions from vehicle sources in 1999–2008')+
  xlab('Year') +
  ylab('Total Emission')

# save plot
 ggsave(file = 'plot5.png')
```

## Compare emissions from motor vehicle sources in Baltimore City with emissions from motor vehicle sources in Los Angeles County, California (fips == "06037"). Which city has seen greater changes over time in motor vehicle emissions?

```{r}
# filter data for Baltimore and Los Angeles 
bl_la <- filter(data, fips == '24510' | fips == '06037' )
table(bl_la$fips)

# get data for emissions from motor vehicle sources
vehicle <- bl_la[grep('[vV]ehicle',bl_la$SCC.Level.Two),]

# group by year and fips and calculate total emission for each year
by_year_fips <- group_by(vehicle, year, fips)
by_year_fips_total <- summarize(by_year_fips, total_emission = sum(Emissions))

# ungroup dataframe before plotting
by_year_fips_total <- ungroup(by_year_fips_total)
by_year_fips_total

# convert fips into city names
names(by_year_fips_total)[2] <- c("City")
by_year_fips_total$City[by_year_fips_total$City == '06037'] <- "Los Angeles"
by_year_fips_total$City[by_year_fips_total$City == '24510'] <- "Baltimore"

# create plot
ggplot(data = by_year_fips_total , aes(x = year, y = total_emission)) +
 geom_line(aes(color = City), size = 4, stat = 'identity')+
  ggtitle('Emissions from vehicle sources in 1999–2008')+
  xlab('Year') +
  ylab('Total Emission')

# save plot
 ggsave(file = 'plot6.png')
```



